The semantic drift of quotations in blogspace: a case study in short-term cultural evolution

### 2016-07-28 brains-appendix.tex
chore/all: typos, colors, todos

\begin{figure*}[!th]
    \centering
    \includegraphics[width=\textwidth]{images/computed-figures/paper-susceptibilities-sentencequantiles.png}
    \caption{\todo{Bring back to 1, and use log-y}}
    \label{fig:feature-susceptibilities-in_sentence}
\end{figure*}

>>>>

\begin{figure*}[!th]
    \centering
    \includegraphics[width=\textwidth]{images/computed-figures/paper-susceptibilities-sentencequantiles.png}
    \caption{\newtext{\textbf{Substitution susceptibility for in-quote feature quartiles:}
    susceptibility to substitution versus quartile of the feature distribution in the originating quote, with 95\% asymptotic confidence intervals (Goodman-based multinomial)}.
    \todo{Bring back to 1, and use log-y}}
    \label{fig:feature-susceptibilities-in_sentence}
\end{figure*}

### 2016-07-28 brains-conclusion.tex
chore/all: typos, colors, todos

>>>>

- place back into the initial question: in our example, do reps converge?
-> we don't know because we couldn't test (ref. sub models and complicated data), but there is attraction.
  - Remind the details
  - It's consistent with known effects (word frequency: easier recall for high freq, age of acquistion: Zevin, Dewhurst, Morrison, clustering coefficient: nelson, age of acquisition (corr pagerank): griffiths, orth. neighborhoos: Garlock), which is interesting, because it's applying cogsci tools on real life data
  - it's also a first step towards analyzing together cogsci and culture (because cogsci on real life data), which is much of a challenge still, although we don't get to see any global effect on the corpus nor reciprocity, therefore (i.e. we don't loop the loop yet)
  - there's a cost to this however: noise, uncertainty (hence single sub), data shaped oppositely to lab experiments.
  - so naturally it's a far cry from the whole story. It's not even the main transformation at work, so we can't conclude on the evolution of the dataset. But we know something more for this simple case, and what it costs to reach that knowledge.

- Now we also looked at context, but there's vast oceans to explore in this direction too. Indeed there's way more to do on this case:
  - in lab exps you can predict the word if proper measures or well-designed lists
  - you can ask what triggers the substitution with semantic similarity: low cost of switching to a more frequent dyad, given the sentence, following Zaromb
  - you can see how previous context influences the semantic similarity (PLI, LSA/LDA). Note that these LSA/LDA methods are ill-suited to the data.
  - with better data, building on our software, we could do that, and look at chains

More broadly with Sperber:
  - you can have a simplistic interpretation, or you can look for attraction on different dimensions, which a more rich (and useful) interpretation.
  - kirby and co have looked at that on artificial languages, on non-meaningful patterns (claidiere), we could look at that on real language.
  - What we see is consistent with lineage specificity (contraction around sentence median), and many things push towards that (context that influences the fitness, as in "language as a system" from kirby, previous history that changes bias, as in PLIs), but right now we don't know in detail what pressure creates that. It's likely the type of task influences that.
  - What we see is not clear w.r.t. convergence however: (1) we don't observe it on our data, and (2) if there's lineage specificity there's no reason convergence should appear (on contrary). But, this is only half the story, since in reality there's also selection, i.e. not all chains go on forever, so you lose some diversity, and it changes the dynamics.
  - finally, context is addressed with relevance in Sperber, but it's still all representational, as we did here. Others have said that it will never be enough. Ingold, Di Paolo.

### 2016-07-28 brains-introduction.tex
chore/all: typos, colors, todos

The theory makes an additional strong hypothesis, which this paper focuses on:
as transformations accumulate, some representations evolve to be very stable and spread throughout an entire society without changing any more (they are called "cultural representations", because they characterize a given culture).
This process should manifest itself as attractors (called "cultural attractors") in the dynamical system that models cultural evolution, that is:
there should be areas of the representation space where cognitive effects in transformations bring representation closer to a given stable asymptotic point.\footnote{
\newtext{Attractors need not be points in fact, they can also be sub-areas; in that case any transformation brings representations in the area closer to (or maintained inside) the target sub-area.
}}

@@

Our question is as follows: given such representations that seem to evolve precisely because of the kind of automatic cognitive biases referred to in the theory of epidemiology of representations, do cultural attractors appear and how do cognitive biases participate in them?
We chose to restrict our analysis to substitutions (one word being replaced by another), both to keep the analysis tractable and because of missing information in our data set.\footnote{
\newtext{As explained further down, source-destination links between quotes must be inferred from the data set, an operation which is much more reliable if we restrict our analysis to substitutions.
This also impedes us from observing the effect of accumulated transformations in the long term, limiting our results to a view of the individual evolutionary step.
}}
While this limits the scope of our observations to the particular data set we use, the methodological point we also make is left intact.
By characterizing words using 6 well-known features, we identify what makes a substitution more likely, and how a word changes when it is substituted.
This exploratory approach uncovers a number of transmission biases consistent with known effects in linguistics.
While the transformations we describe are not the only ones at work in this data set, our analysis also indicates that feature-specific attractors could exist because of the substitution process.
This study can be viewed as analyzing part of the transmission step operating in transmission chains of artificial languages like those studied by \citet{kirby_cumulative_2008}, but with natural language out of the laboratory.

@@

Compared to full sentences, recall of word lists provides a situation that is easier to fully explore, and that has been extensively studied.
In particular, the Deese, Roediger, and McDermott paradigm (introduced by \citealp{deese_prediction_1959}, and later popularized by \citealp{roediger_creating_1995}) has shown that it is possible to construct lists of words which reliably create the false memory of an external word related to those in the list.
This is done by using lists of words produced by free association from the target intrusion word;
the intruding recall then happens with probability nearly proportional to the average strength of semantic association between the intruding word and the words in the list.
A sizable literature studies this type of task with varying complexities in the design of the lists, a good review of which is given by \citet{zaromb_temporal_2006}.
One notable effect is that the semantic relations between words greatly influence, and correlate to, the order in which words are recalled~\citep{tulving_subjective_1962,howard_when_2002}, and that this reordering of items improves subjects' repeated recalls~\citep{tulving_subjective_1966}.
The frequency and type of intrusions in lists of random words are also influenced by associations created by the presentation of previous lists~\citep{zaromb_temporal_2006};
indeed the question of how such temporal associations (contributing to contextual information retrieval in recall) interact with the prior semantic associations of subjects (contributing to associative information retrieval) is at the center of many of these studies.

@@

Given this complexity we decided to focus on more aggregate effects, where variations of the conditions in which sentences are read and produced have a chance of being statistically smoothed out.\footnote{
\newtext{Aside from our lack of control on the precise conditions of encoding and recall in our data set, the analysis techniques mentioned above are better suited to data consisting of a high number of measures over a smaller number of lists (in which case it makes sense to ask \hbox{e.g.} what proportion of intrusions come from prior lists).
As is explained further down however, our data set is shaped the opposite way: a great number of sentences, with only very few to no measures at all on each sentence.
}}
If a cognitive bias in the substitution of words manifests itself with simple measures, then it will be worth applying predictive models of the substitution process in further research.

>>>>

The theory makes an additional strong hypothesis, which this paper focuses on:
as transformations accumulate, some representations evolve to be very stable and spread throughout an entire society without changing any more (they are called "cultural representations", because they characterize a given culture).
This process should manifest itself as attractors (called "cultural attractors") in the dynamical system that models cultural evolution, that is:
there should be areas of the representation space where cognitive effects in transformations bring representations closer to a given stable asymptotic point.\footnote{
\newtext{Attractors need not be points in fact, they can also be sub-areas; in that case any transformation brings representations in the area closer to (or maintained inside) the target sub-area.
}}

@@

Our question is as follows: given such representations that seem to evolve precisely because of the kind of automatic cognitive biases referred to in the theory of epidemiology of representations, do cultural attractors appear and how do cognitive biases participate in them?
We chose to restrict our analysis to substitutions (one word being replaced by another), both to keep the analysis tractable and because of missing information in our data set.\footnote{
\newtext{As explained further down, source-destination links between quotes must be inferred from the data set, an operation which is much more reliable if we restrict our analysis to substitutions.
This also impedes us from observing the effect of accumulated transformations in the long term, limiting our results to a view of the individual evolutionary step.
}}
While this limits the scope of our results to the particular data set we use, the methodological point we also make is left intact.
By characterizing words using 6 well-studied features, we identify what makes a substitution more likely, and how a word changes when it is substituted.
This exploratory approach uncovers a number of transmission biases consistent with known effects in linguistics.
While the transformations we describe are not the only ones at work in this data set, our analysis also indicates that feature-specific attractors could exist because of the substitution process.
This study can be viewed as analyzing part of the transmission step operating in transmission chains of artificial languages like those studied by \citet{kirby_cumulative_2008}, but with natural language out of the laboratory.

@@

Compared to full sentences, recall of word lists provides a situation that is easier to fully explore, and that has been extensively studied.
In particular, the Deese, Roediger, and McDermott paradigm (introduced by \citealp{deese_prediction_1959}, and later popularized by \citealp{roediger_creating_1995}) has shown that it is possible to construct lists of words which reliably create the false memory of an external word related to those in the list.
This is done by using lists of words produced by free association from the target intrusion word;
the intruding recall then happens with probability nearly proportional to the average semantic association strength between the intruding word and the words in the list.
A sizable literature studies this type of task with varying complexities in the design of the lists, a good review of which is given by \citet{zaromb_temporal_2006}.
One notable effect is that the semantic relations between words greatly influence, and correlate to, the order in which words are recalled~\citep{tulving_subjective_1962,howard_when_2002}, and that this reordering of items improves subjects' repeated recalls~\citep{tulving_subjective_1966}.
The frequency and type of intrusions in lists of random words are also influenced by associations created by the presentation of previous lists~\citep{zaromb_temporal_2006};
indeed the question of how such temporal associations (contributing to contextual information retrieval in recall) interact with the prior semantic associations of subjects (contributing to associative information retrieval) is at the center of many of these studies.

@@

\newtext{Given this complexity we decided to focus on more aggregate effects, where variations of the conditions in which sentences are read and produced have a chance of being statistically smoothed out.\footnote{
\newtext{Aside from our lack of control on the precise conditions of encoding and recall in our data set, the analysis techniques mentioned above are better suited to data consisting of a high number of measures over a smaller number of lists (in which case it makes sense to ask \hbox{e.g.} what proportion of intrusions come from prior lists).
As is explained further down however, our data set is shaped the opposite way: a great number of sentences, with only very few to no measures at all on each sentence.
}}
If a cognitive bias in the} substitution \newtext{of words manifests itself with simple measures, then it will be worth applying predictive models of the substitution process in further research.}
